---
title: "Case Study 1 - Part 1 - Attrition EDA"
author: "Robert Hines"
date: "2024-06-30"
output: word_document
---

### Youtube Video Link https://www.youtube.com/watch?v=6Pa3KNNiNt4 ###

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Load the data
attrition <- read.csv("PATH/CaseStudy1-data.csv")

```

```{r}
ggplot(attrition, aes(x = MonthlyIncome, y = JobSatisfaction, color = Attrition)) +
  geom_point(alpha = 0.7) +
  labs(title = "Monthly Income vs. Job Satisfaction by Attrition", x = "Monthly Income", y = "Job Satisfaction") +
  facet_wrap(~ Attrition) +
  theme(legend.position = "none")


ggplot(attrition, aes(x = Age, fill = Attrition)) +
  geom_histogram(binwidth = 1, position = "dodge") +
  labs(title = "Age and Attrition by Job Role", x = "Age", y = "Count") +
  facet_wrap(~ JobRole) +
  theme(legend.position = "none")


ggplot(attrition, aes(x = TotalWorkingYears, fill = Attrition)) +
  geom_histogram(binwidth = 1, position = "dodge") +
  labs(title = "Total Working Years and Attrition by Job Role", x = "Total Working Years", y = "Count") +
  facet_wrap(~ JobRole) +
  theme(legend.position = "none")


# Visualize employee turnover by JobRole (assuming there is a Turnover column with 1 for turnover and 0 for retained)
ggplot(attrition, aes(x = JobRole, fill = factor(Attrition))) +
  geom_bar(position = "fill") +
  labs(title = "Employee Turnover by Job Role",
       x = "Job Role",
       y = "Proportion",
       fill = "Turnover") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_fill_manual(values = c("Yes" = "#FFA07A", "No" = "#87CEFA"))  # Custom colors
```

```{r}
# Perform Kruskal-Wallis test
kruskal_test <- kruskal.test(Attrition ~ JobRole, data = attrition)
kruskal_test
```
```{r}
# Perform Kruskal-Wallis test
kruskal_test <- kruskal.test(JobRole ~ Attrition, data = attrition)
kruskal_test
```

```{r}
attrition_by_gender_role <- attrition %>%
  group_by(JobRole, Gender, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

ggplot(attrition, aes(x = WorkLifeBalance, y = JobSatisfaction, fill = Attrition)) +
  geom_boxplot() +
  labs(title = "Work-Life Balance vs. Job Satisfaction by Attrition", x = "Work-Life Balance", y = "Job Satisfaction") +
  facet_wrap(~ Attrition) +
  theme(legend.position = "none")

ggplot(attrition, aes(x = YearsAtCompany, y = PerformanceRating, color = Attrition)) +
  geom_jitter(alpha = 0.7) +
  labs(title = "Years at Company vs. Performance Rating by Attrition", x = "Years at Company", y = "Performance Rating") +
  facet_wrap(~ Attrition) +
  theme(legend.position = "none")

attrition_by_department_role <- attrition %>%
  group_by(Department, JobRole, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

ggplot(attrition_by_department_role, aes(x = Percentage, y = JobRole, fill = Attrition)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Attrition by Department and Job Role", x = "Percentage", y = "Job Role") +
  facet_wrap(~ Department) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("Yes" = "#FFA07A", "No" = "#87aFFF"))

ggplot(attrition_by_gender_role, aes(x = Percentage, y = JobRole, fill = Attrition)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Attrition by Gender and Job Role", x = "Percentage", y = "Job Role") +
  facet_wrap(~ Gender) +
  scale_fill_manual(values = c("Yes" = "#FFA07A", "No" = "#87aFFF"))

gender_by_department_role <- attrition %>%
  group_by(Department, JobRole, Gender) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

ggplot(gender_by_department_role, aes(x = Percentage, y = JobRole, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Gender by Department and Job Role", x = "Percentage", y = "Job Role") +
  facet_wrap(~ Department)

```

```{r}
attrition_by_education <- attrition %>%
  group_by(EducationField, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

ggplot(attrition_by_education, aes(x = EducationField, y = Percentage, fill = Attrition)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Attrition Rates by Education Field", x = "Education Field", y = "Percentage") +
  scale_fill_manual(values = c("Yes" = "#FFA07A", "No" = "#87aFFF"))
```
```{r}
# Perform Kruskal-Wallis test
kruskal_test <- kruskal.test(Attrition ~ EducationField, data = attrition_by_education)
kruskal_test
```


```{r}
ggplot(attrition, aes(x = MonthlyIncome, y = EducationField, color = Attrition)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Monthly Income vs. Education Field by Attrition", x = "Monthly Income", y = "Education Field") +
  theme(legend.position = "none") +
  scale_color_manual(values = c("Yes" = "#FFA07A", "No" = "#87aFFF"))

ggplot(attrition, aes(x = MonthlyIncome, y = JobRole, color = Attrition)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Monthly Income vs. Job Role by Attrition", x = "Monthly Income", y = "Job Role") +
  theme(legend.position = "none") +
  scale_color_manual(values = c("Yes" = "#FFA07A", "No" = "#87aFFF"))

ggplot(attrition, aes(x = MonthlyIncome, y = Department, color = Attrition)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Monthly Income vs. Job Role by Attrition", x = "Monthly Income", y = "Job Role") +
  theme(legend.position = "none") +
  scale_color_manual(values = c("Yes" = "#FFA07A", "No" = "#87aFFF"))
```


```{r}
attrition_by_education_gender <- attrition %>%
  group_by(EducationField, Gender, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

ggplot(attrition_by_education_gender, aes(x = EducationField, y = Percentage, fill = Attrition)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Attrition by Education Field and Gender", x = "Education Field", y = "Percentage") +
  facet_wrap(~ Gender) +
  theme(legend.position = "none")

```

```{r}
attrition_by_role_gender <- attrition %>%
  group_by(JobRole, Gender, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

ggplot(attrition_by_role_gender, aes(x = JobRole, y = Percentage, fill = Attrition)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Attrition by Job Role and Gender", x = "Job Role", y = "Percentage") +
  facet_wrap(~ Gender) +
  theme(legend.position = "none")
```

```{r}
# Convert Attrition to a factor
attrition$Attrition <- as.factor(attrition$Attrition)

# Perform ANOVA
anova_model <- aov(Percentage ~ JobRole * Gender, data = attrition_by_role_gender)
summary(anova_model)
```

JobRole:
F value: 0.038
Pr(>F): 1.000
The extremely high p-value suggests that job role has no statistically significant effect on attrition percentage.

Gender:
F value: 0.052
Pr(>F): 0.823
The high p-value indicates that gender does not have a statistically significant effect on attrition percentage.

JobRole:Gender
F value: 0.077
Pr(>F): 1.000
The very high p-value suggests that the interaction between job role and gender is not statistically significant in affecting attrition percentage.


```{r}
# Perform Kruskal-Wallis test for JobRole
kruskal_jobrole <- kruskal.test(Percentage ~ JobRole, data = attrition_by_role_gender)
kruskal_jobrole

# Perform Kruskal-Wallis test for Gender
kruskal_gender <- kruskal.test(Percentage ~ Gender, data = attrition_by_role_gender)
kruskal_gender
```

```{r}
ggplot(attrition, aes(x = WorkLifeBalance, y = JobSatisfaction, color = Attrition)) +
  geom_jitter(alpha = 0.7) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Work-Life Balance vs. Job Satisfaction by Job Role", x = "Work-Life Balance", y = "Job Satisfaction") +
  facet_wrap(~ JobRole) +
  theme(legend.position = "none")
```

```{r}
ggplot(attrition, aes(x = MonthlyIncome, y = OverTime, color = Attrition)) +
  geom_jitter(width = 0.3, height = 0.3) +
  theme(axis.text.x = element_text(hjust = 1)) +
  labs(title = "Attrition Rates based on Monthly Income vs. Overtime by Job Role", x = "Monthly Income", y = "Overtime") +
  facet_wrap(~ JobRole) +
  scale_color_manual(values = c("Yes" = "#FFA07A", "No" = "#87aFFF")) +
  theme(legend.position = "none")
```

```{r}
# Fit a linear model
model <- lm(MonthlyIncome ~ OverTime * JobRole, data = attrition)

# Summary of the model to check significant interactions and main effects
summary(model)

# Optionally, you can use ANOVA to test the model
anova(model)

```
